---
title: "Development Analysis of Austria, Latvia, and Poland"
subtitle: "Economic, Social, Environmental and Health Indicators"
author: "Vimarish K M | 24229318"
date: "10 June 2025"
format:
  beamer:
    toc: false
    template: header_beamer.tex
    #theme: metropolis
    #colortheme: beaver
    #slide-level: 2
    fig-align: center
fontsize: 10pt
execute:
  echo: true
  message: false
  warning: false
  size: 15
---

## Objective

-   Compare economic, social, and development indicators across three countries.

-   Focus: **data.table** speed and efficiency.

-   Countries: Austria, Latvia, Poland.

-   Data Source: World Bank (HDX).

## Load Packages

```{r}
#| warning: false
library(data.table)
library(ggplot2)
library(stringr)
library(dplyr)
library(knitr)
library(latex2exp)
library(readr)
library(plotly)
library(tidyr)
```

## Read and Clean Data

```{r}
# Read the data (skipping metadata row)
correct_names = c("Country Name", "Country ISO3",
                  "Year","Indicator Name",
                  "Indicator Code", "Value")
aut = fread("indicators_aut.csv", skip = 2,
            col.names = correct_names)
lva = fread("indicators_lva.csv", skip = 2,
            col.names = correct_names)
pol = fread("indicators_pol.csv", skip = 2,
            col.names = correct_names)
# Merge all datasets into one
data_all = rbindlist(list(aut, lva, pol))
# Assign correct classes
data_all[, Year := as.integer(Year)] # Year to integer  
data_all[, Value := as.numeric(Value)] # Value to num  
```

## Quick Data Exploration

```{r}
#| label: data-summary
#| echo: true
#| eval: true
# 1. Dimensions and structure
str(data_all, vec.len = 1.4)
```

## Summary:

The combined data set comprises 224,813 observations across 6 variables. The key fields include:

-   **Country Name**, **Country ISO3**, **Indicator Name**, and **Indicator Code**: stored as character variables.

-   **Year**: correctly recognized as an integer.

-   **Value**: stored as a numeric variable representing the actual measurement for each indicator.

```{r}
#| label: unique-checks
#| echo: true
#| eval: true
kable(data_all[, .(Indicators = uniqueN(`Indicator Name`),
             Years = paste0(min(Year), "-", max(Year))), 
         keyby = `Country Name`])
```

## Interpretation

The dataset includes indicator data for the following countries:

-   **Austria** has data for **1,998 indicators** spanning the years **1960 to 2024**.

-   **Latvia** has data for **3,161 indicators** over the same period (**1960 to 2024**).

-   **Poland** has the most extensive coverage, with **3,584 indicators** from **1960 to 2024**.

-   For the Indicator in terms of range of years, we group by country using "keyby".

## Country Count

```{r}
#| label: count-by-country
#| echo: true
#| eval: true
kable(data_all[, .(Total_Observations = .N),
         keyby = `Country Name`])
```

-   Austria has 73,499 records, Latvia has 64,771 records, and Poland has 86,543 records in the data set.

-   All three countries are well-represented, with no major imbalance in data volume. This balanced representation supports a fair and reliable comparative analysis across countries.

## Indicator Frequency Table

```{r}
#| label: Indicator Frequency
#| echo: true
#| eval: true

kable(data_all[, .N,
               keyby = `Indicator Name`][order(-N)][1:10])
```

## Top Indicators (by Mean Value)

```{r}
#| label: indicator-stats
#| echo: true
#| eval: true
#| fig-align: center

# Summary of combined data by mean value
ind_summary = data_all %>%
  group_by(`Indicator Name`) %>%
  summarise(
    #min = min(Value, na.rm = TRUE),
    mean_value = mean(Value, na.rm = TRUE),
    #max = max(Value, na.rm = TRUE),
    Total = n()
  ) %>% 
  arrange(desc(mean_value)) %>%
  select(-mean_value)
```

## 

```{r}
print(ind_summary,n = 10)
```

## Interpretation

The top indicators, based on average values across years and countries, include:

1.  GNI (constant LCU)
2.  Broad money (current LCU)
3.  Gross Domestic Income (constant LCU)
4.  Gross national expenditure (constant LCU)
5.  GNI, PPP (constant 2021 international \$ )

These are primarily **economic indicators** with high monetary values, mostly expressed in local currency units (LCU). This suggests that the data set offers rich coverage of **macroeconomic metrics**, making it well-suited for analyzing national income, expenditure, and monetary trends over time.

## Electricity Access (%)

```{r}
#| label: Electricity-Access
#| echo: true
#| eval: true

# Electricity access by country
elec = data_all[`Indicator Name` ==
    "Access to electricity (% of population)"]

elec[, .(MeanAccess = mean(Value, na.rm = TRUE)),
     keyby = `Country Name`]
```

## Life Expectancy Over Years by Country

```{r}
#| warning: false  
#| echo: false
#| eval: true
#| fig-align: center
#| out-width: 90%
 
# Exploring Life Expectancy by Country
life_exp = data_all[`Indicator Name` ==
    "Life expectancy at birth, total (years)"]

ggplot(life_exp, aes(x = Year, y = Value,
                     color = `Country Name`)) +
  geom_line(size = 1.2) + theme_minimal() +
  labs(title = "Life Expectancy Over Time")

```

## Interpretation

The chart shows a clear upward trajectory in life expectancy across all three countries (Austria, Latvia, and Poland) from 1960 to 2020, reflecting global improvements in healthcare and living standards.

-   **Austria** maintains the highest life expectancy throughout the period (reaching \~80 years by 2020)

-   **Latvia** and **Poland** show parallel growth patterns, with Latvia slightly ahead in recent years

-   A noticeable dip occurs around 1990-2000 for Latvia/Poland, likely reflecting post-Soviet transition challenges

<!-- -->

-   The most rapid improvements occurred between 1960-1980

-   Convergence trend appears post-2000 as Latvia/Poland narrow the gap with Austria

## GDP per Capita

```{r}
#| warning: false  
#| echo: false
#| eval: true
#| fig-align: center
#| out-width: 90%

# Filter GDP per Capita by country
gdp = data_all[`Indicator Name` ==
    "GDP per capita (current US$)"]
ggplot(gdp, aes(x = Year, y = Value,
                color = `Country Name`)) +
  geom_line() + theme_minimal() +
  labs(title = "GDP per Capita (US$)")
```

## Interpretation

The chart shows **steady growth in GDP per capita** from 1980 to 2000, indicating economic expansion over two decades.

-   GDP per capita **doubled** from \~\$20,000 (1980) to \~\$40,000 (2000), reflecting significant improvements in productivity or income distribution.

-   A sharp rise occurs after 2000, driven largely by Austria high GDP values.

-   Fluctuations post-2008 reflect the global financial crisis, and the dips around 2020 likely capture the COVID-19 economic impact for Austria where it doesn't affect much for Latvia and Poland.

# ***Thank You***
